---
title: "final_report"
output: html_document
date: "2022-12-13"
---
## Differential gene expression analysis using Deeq2

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

```

### Loading the libraries 

```{r a,,echo=TRUE}
library("DESeq2")
library(limma)
library(edgeR) 
library(RColorBrewer)
library(clusterProfiler)
library(enrichplot)
library(ggplot2)
library(org.Hs.eg.db)
library(Glimma)
```

### Load the data

```{r b,,echo=TRUE}
data <- as.matrix(read.csv(file = 'X:/fast/gilbert_p/fg_data/SCRI/TBVPX-203/RNA/2019Dec/salmon_merged_gene_counts.csv', row.names = "gene_id"))
colnames(data) = gsub("X", "P", colnames(data))
```

### Input treatment data
```{r c,echo=TRUE}
trt_data <- read.csv('trt.csv', header=FALSE, stringsAsFactors=FALSE, fileEncoding="latin1")
colnames(trt_data) = trt_data[1,1:5]
trt_data=trt_data[-1,]
trt_data<-trt_data[trt_data$`Treatment Group` %in% c("2 µg ID93 + 5 µg GLA-SE (2 Vaccine Injections)", "2 µg ID93 + 5 µg GLA-SE (3 Vaccine Injections)"), ]
head(trt_data)
```

### Input sample meta data
```{r d,echo=TRUE}
samples <- read.csv('X:/fast/gilbert_p/fg_data/SCRI/TBVPX-203/RNA/2019Dec/salmon_metadata.csv')
samples<-samples[,2:4]
rownames(samples) <- samples$samplename
samples_0_3<-samples
samples_0_3$Subject_ID<-substr(samples_0_3$samplename,1,(tail(unlist(gregexpr('_', samples_0_3$samplename)), n=1))-1)
samples_0_3$Subject_ID =gsub("P203_", "", samples_0_3$Subject_ID)
samples_0_3$Subject_ID=gsub("_", "", samples_0_3$Subject_ID)
samples_0_3<-samples_0_3[samples_0_3$Subject_ID %in% trt_data$`Subject ID`, ]
tt<-merge(trt_data,samples_0_3, by.x= "Subject ID", by.y="Subject_ID")
head(tt)
```


### Subset data for Deseq2
```{r e,echo=TRUE}
tt_0_3<-tt[tt$day %in% c("0", "3"), ]
tt_0_3$day <- factor(tt_0_3$day)
rownames(tt_0_3)<-tt_0_3$samplename

idx <- match(rownames(tt_0_3), colnames(data))
data_0_3<- data[,idx]
all(rownames(tt_0_3) == colnames(data_0_3))

```


### Create a DESEQ object
```{r f,echo=TRUE}

dds <- DESeqDataSetFromMatrix(countData = round(data_0_3),
                              colData = tt_0_3,
                              design= ~ ptid+day)

```



### Filter genes for Deseq2
```{r g,echo=TRUE}

keep <- rowSums(counts(dds) >= 10) >= 10
sum(keep)
dds <- dds[keep,]
nrow(dds)

```

### Estimate dispersion in DEseq object
```{r h,echo=TRUE}

dds <- estimateSizeFactors(dds)
sizeFactors(dds)
dds = estimateDispersions( dds )

```



### Plot dispersion estimate
```{r i,echo=TRUE}

plotDispEsts(dds)
cds <- DESeq(dds)
res <- results(cds)
table(res$padj<0.1)
```





### Find out significant genes


```{r j,echo=TRUE}

up = res[ which(res$padj < 0.1 & res$log2FoldChange > 0), ]
down = res[ which(res$padj < 0.1 & res$log2FoldChange < 0), ]
signicant_genes = rbind(up, down)
rownames(up)
rownames(down)
```




### Order the result from deseq based on pvalues
```{r k,echo=TRUE}

res <- res[order(res$padj), ]
head(res)
```


### Merge with normalized count data
```{r l,echo=TRUE}

result_p_value <- merge(as.data.frame(res), 
                 as.data.frame(counts(dds, normalized = TRUE)), 
                 by = "row.names", 
                 sort = FALSE)
names(result_p_value)[1] <- "Gene"
head(result_p_value)
```

### Write results

```{r n,echo=TRUE}
write.csv(result_p_value, file = "Deseq2_diffexpr-results_3_vs_0.csv", quote = FALSE, row.names = F)

write.csv(signicant_genes, file = "Deseq2_significant_genes-results_3_vs_0.csv", quote = FALSE, row.names = T)
```

